\ripple: Differentiable Waveforms for Gravitational Wave Data Analysis

### 2022-12-15 src/tex/ms.tex
Started adding text to HMC and first part of discussion

\section{Discussion}
\label{subsec:discussion}

>>>>

After the search algorithms have constructed a list of confidently detected binaries, the next step is to sample from the posterior of each sources parameters - so called parameter estimation (PE).
To do this, one typically uses a Markov Chain Monte Carlo (MCMC) sampler or nested sampler~\citep{multinest, dynesty}.
Although robust, both MCMC and nested sampling are slow to converge and are known to perform poorly in high dimensional parameter spaces.
For example, sampling the 15 dimensional parameter space for a BBH system can take $\mathcal{O}(10)$hours, whilst BNS systems can take up to weeks.
Dedicated fast samplers have been designed to get approximate posteriors on the sky localization to faciliate follow-up electromagnetic observations.
Nevertheless, these do not present the whole picture; fast, general parameter estimation therefore remains a primary goal of GW data analysis.

A primary issue with both MCMC and nested sampling is that neither utlizes information about the likelihood's derivative and must therefore randomly walk towards areas of highest likelihood.
Derivative based samplers, on the other hand, have been shown to extrapolate well to higher dimensions although they sometimes come with their own drawbacks.
Here we simply aim to demonstrate the utility of a derivative based sampler and their efficiency on a small test problem.
In particular, we will show that proposals made using an Hamiltonian Monte Carlo (MC) sampler are accepted at a much higher rate than a traditional MCMC algorithm.
In a follow up paper we will demonstrate that minute scale PE can be achieved by exploiting normalizing flows~\citep{2022arXiv221106397W, Gabrie:2021tlu} together with a derivative based sampler~\citep{PEpaper}.

\section{Discussion and Conclusion}
\label{subsec:discussion}

In this paper we introduced and discussed the various benefits of differentiable waveforms in \jax for GW data analysis.
First, we demonstrated the speed and accuracy of our implementation of the aligned spin IMRPhenomD waveform.
In particular, we showed that it matches the \lalsuite implementation to near machine precision and can be easily parallelized on a GPU, making waveform calls up to five orders of magnitude faster.
Second, we discussed three data analysis tasks which can all be substantially improved by utlizing derivative information of the waveform.
Although we primarily discuss toy examples in this paper, each can be extended to the full data analysis task. 
Differentiable waveforms therefore represent a crucial tool to perform efficient GW science.

Currently the biggest constraint to adopting differentiable waveforms is the need to re-write the most commonly used waveforms into \jax.
At the time of writing...
